MLExplain

### 2018-05-24 paper.tex
done with JSExplain description

\begin{abstract}
  MLExplain is a step-by-step interpreter for OCaml that enables the
  user to inspect both their program's state and the interpreter's state
  itself. This interpreter aims to provide the user with a better understanding
  of the semantics of OCaml semantics.
\end{abstract}

@@

The goal of the \emph{JSExplain} project \cite{chargueraud:hal-01745792} is to
describe the execution of a JavaScript program by showing every step of an
interpreter whose behavior is very closest to the specification of JavaScript.
In this paper, we show how we adapted JSExplain to OCaml.

Unlike JavaScript, OCaml has no official specification. However, OCaml code
execution is fairly simple because the amount of language constructions is
small. We have written an interpreter for OCaml's typed abstract syntax tree
(AST). This AST is close to the source code and need little transformations but
supplies us with resolved names -- which is useful for module and signatures
inclusion and absolutely mandatory for features like named parameters and
optional parameters in functions. The semantic we give to OCaml is higher level
than the one described in \emph{ZINC} \cite{Leroy-ZINC} virtual machine which is
at the same level as bytecode.

@@

\subsection{Double Debugger}
\label{subsec:double-debugger}

@@

ALAN: stopped here

JSExplain is a visual interpreter. To display its state, the interpreter is
compiled with a compiler that adds calls to a tracing function. The traced
retrieved during the execution of the program gives many information about the
state of the interpreter:

\begin{itemize}
  \item existing variables and their values
  \item return values of functions of the interpreter
\end{itemize}

JSExplain's web interface uses these traces in order to display the program and
interpreter state at every step of the execution.

The whole project takes the shape of a web application composed of the
JavaScript interpreter and a view. The view allows the user to write JavaScript
code and run it. It also displays the state of the program and the state of the
interpreter.

The program's execution is not interactive: when the button \texttt{run} is
pressed, the program is executed all at one and the traces are generated. Only
then, the user can replay the execution step by step.

\subsection{JSExplain's compiler}

% Référence pour coq ?
JSExplain was first written in Coq. OCaml code has been extracted from the Coq
source and is now the primary source code. The interpreter is compiled into
JavaScript in order to use it within a web page. For this reason, the project
embeds a homemade OCaml-to-JavaScript compiler.

JSExplain also needs its own OCaml to JavaScript compiler in order to place the
calls to \verb|log_event| which is the tracing function. This function is also
used to get the location of the different syntactical objects in the source
file in order to produce code highlighting in JSExplain's interface.

This compiler offers several compilation modes:

\begin{itemize}
  \item \texttt{unlog}: generate JavaScript without tracing.
  \item \texttt{log}: generate JavaScript with tracing.
  \item \texttt{mlloc}, \texttt{token} and \texttt{ptoken}: generate files with
    informations about syntastical elements from the input file.
  \item \texttt{pseudo}: generate JavaScript enhanced with syntactic sugar for
    pattern matching or monadic operations -- it is supposed to be easier to
    read.
\end{itemize}

Modes \texttt{unlog} and \texttt{log} exist so that auxiliary code in the
interpreter -- like linked-list or hash-map internal-management functions -- is
not traced.  The user is thus able to concentrate on the logical part of the
interpreter functioning and forget about implementation details.

\subsubsection{fOCaml}
This compiler does not recognized the whole OCaml language, it can only handle
a small subset we named \texttt{fOCaml}. Among other restrictions, it cannot
handle objects and imperative features, exceptions, inner modules, and tuples.
The main reason for this limitation is that compiling OCaml into JavaScript is
not trivial, many features -- e.g. nested pattern matching -- are hard to
represent to JavaScript.

The fOCaml code is compiled into \texttt{fJs}, a functional subset of
JavaScript. The compiler preserves as much as possible the code structure to
enable the user to understand both fOCaml and fJs codes as the same one.

% Petite explication sur le fait que du coup on écrit du code fonctionnel pur
% et les implications que ça a ?

>>>>

\begin{abstract}
  MLExplain is a step-by-step interpreter for OCaml that enables the
  user to inspect both their program's state and the interpreter's state
  itself. This interpreter aims to provide the user with a better understanding
  of the semantics of OCaml.
\end{abstract}

@@

The goal of the \emph{JSExplain} project \cite{chargueraud:hal-01745792} is to
describe the execution of a JavaScript program by showing every step of an
interpreter whose behavior is very close to the specification of JavaScript. In
this paper, we show how we adapted JSExplain to OCaml.

Unlike JavaScript, OCaml has no official specification. However, OCaml code
execution is fairly simple because the number of language constructions is
small. We have written an interpreter for OCaml's typed abstract syntax tree
(AST). This AST is still close to the source code yet it provides additional
crucial information. For instance, it mentions resolved names, which is useful
for module and signatures inclusion and mandatory for interpreting features like
named parameters and optional parameters in functions. The semantic we give to
OCaml is higher level than the one described in \emph{ZINC} \cite{Leroy-ZINC}
virtual machine, i.e., an interpreter for bytecode after compilation.

@@

\subsection{A Double Debugger}
\label{subsec:double-debugger}

@@

We instrument the compilation to JavaScript by adding code that logs events,
namely entering a function, creating a scope, assigning a variable, and exiting
a function. Each event captures the state, the stack, and the values of all
local variables in scope of the interpreter code at the point where the event
gets triggered. Code that is traced is compiled with this instrumentation,
whereas code that is not traced, for instance supporting libraries, is directly
compiled to JavaScript.

When using the tool, the source JavaScript program is parsed and run through the
instrumented interpreter, thus generating a trace. We can then navigate this
trace to see the state of the interpreter as well as the state of the
interpreted program. Recovering the information about the interpreted code is
not completely straightforward. For example, to recover the fragment of code to
highlight, we find in the trace the closest previous event that contains a call
to function with an argument named \texttt{\_term\_}. This argument corresponds
to the AST of a subexpression, and this AST is decorated (by the parser) with
locations. Note that, for efficiency reasons, we associate to each event from
the trace its corresponding \texttt{\_term\_} argument during a single pass,
performed immediately after the trace is produced.

Similarly, we are able to recover the state and environment associated with the
event. The state of the interpreted program consists of four fields: the
strictness flag, the value of the \texttt{this} keyword, the lexical
environment, and the variable environment. We implemented a custom display for
these elements, and also for values of the languages, in particular for objects:
one may click on an object to reveal its contents and recursively explore it.

Finally, a web interface is presented to the user, to navigate in the trace.
This navigation can be forward and backward, and at all times the source code
and interpreter code are highlighted, and their states are displayed. We also
provide a way to specify breakpoints, using arbitrary JavaScript expressions,
that reference both the interpreter and interpreted programs. One may thus reach
a point where some interpreter line is being evaluated with some source value
satisfying a predicate, or vice-versa.
